---
title: "Visualization"
output: github_document
date: "2026-10-01"
---

```{r setup, include=FALSE}
library(tidyverse)
library(ggridges)
library(p8105.datasets)

data("weather_df")
```

Now we have everything we need! 

```{r}
weather_df
```

Let's make a scatterplot! 

```{r}
ggplot(weather_df, aes(x= tmin, y = tmax)) + geom_point()
```
I always like the dataframe first.

```{r}
ggp_temp_scatterplot = 
weather_df |>
  ggplot(aes(x = tmin, y =tmax)) +
  geom_point()

ggp_temp_scatterplot
```

Let's make this fancier

alpha = .25 sets the transparency/opacity of the points in a geom_point() plot.
```{r}
weather_df |>
  ggplot(aes(x = tmin, y =tmax, color = name)) +
  geom_point(alpha = .25) +
  geom_smooth(se = FALSE)
```

```{r}
weather_df |>
  ggplot(aes(x = tmin, y =tmax)) +
  geom_point(aes(color = name), alpha = .25) +
  geom_smooth(se = FALSE)
```

Do not need aesthetics

```{r}
weather_df |>
  ggplot(aes(x = tmin, y =tmax, color = name)) +
  geom_smooth(se = FALSE)
```

Show faceting

```{r}
weather_df |>
  ggplot(aes(x = tmin, y =tmax, color = name)) +
  geom_point(alpha = .5)+
  facet_grid(. ~ name)
```

This is the same as the previous code chunk, just a different coding convention 

```{r}
weather_df |>
  ggplot(aes(x = tmin, y =tmax, color = name)) +
  geom_point(alpha = .5)+
  facet_grid(cols = vars(name))
```
Let's look at something else. 

```{r}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(alpha = .5) +
  facet_grid(. ~ name)
```


```{r}
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = .5) +
  geom_smooth(se = FALSE) + 
  facet_grid(. ~ name)
```

make a plot of central park tmax v tmin only, and convert temperatures to fahrenheit. 

```{r}
weather_df |>
  filter(name == "CentralPark_NY") |>
  mutate(
    tmax = tmax * (9/5) + 32,
    tmin = tmin * (9/5) + 32
  ) |>
  ggplot(aes(x = tmin, y =tmax)) +
  geom_point()
  
```

What's a hex plot 

```{r}
weather_df |>
  ggplot(aes(x = tmin, y = tmax)) +
  geom_hex()
```

## Univariate plots

```{r}
weather_df |>
  ggplot(aes(x = tmax)) +
  geom_histogram()
```
```{r}
weather_df |>
  ggplot(aes(x= tmax, fill = name)) +
  geom_histogram() +
  facet_grid(. ~ name)
```
Density plots (smoothed out histogram)

```{r}
weather_df |>
  ggplot(aes(x = tmax, color = name)) +
  geom_density(alpha = .3)

```

```{r}
weather_df |>
  ggplot(aes(x = tmax, fill = name)) +
  geom_density(alpha = .3)

```

Boxplots 

```{r}
weather_df |>
  ggplot(aes(x = name, y= tmax)) +
  geom_boxplot()
```

Violin Plots 

```{r}
weather_df |>
  ggplot(aes(x = name, y = tmax)) +
  geom_violin()

```

Ridge plots 

```{r}
weather_df |>
  ggplot(aes(x = tmax, y = name)) +
  geom_density_ridges()
```

## Save some of my plots 

```{r}
ggp_weather = 
weather_df |>
  ggplot(aes(x = date, y = tmax, color = name)) +
  geom_point(aes(size = prcp), alpha = .5) +
  facet_grid(. ~ name)
  
ggp_weather

ggsave("ggp_weather.pdf", ggp_weather)
```

```{r fig.width = 12, fig.height = 12}
weather_df |>
  ggplot(aes(x = tmin, y = tmax))+
  geom_point()
```

